Ownership types in theory and practice (in Rust)

### 2023-01-20 paper/paper.tex
Revamped transition from theory to rust

\keywords{Ownership, Type, Safety, Rust}

\maketitle

@@

% TODO: define memory safety
% Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.

@@

We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
% TODO: change wording from source
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
The main problem here is that such memory issues cannot be detected at compile time.

@@

% A quick definition of ownership: you cannot have mutability and alias at the same time.
% My own example:

@@

% TODO: find better disadvantage
A disadvantage of the model may be the added complexity of read-only references.
However, its advantage is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.

% \subsection{Limitations}
% \subsection{From theory to Rust}

% These early ownership type models carry some flaws with them.
% For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.

% TODO: chapter 6 of ownership-types-survey

@@

Rust's notion of ownership is the culmination of a long line of work.
Wadler [1991] and Baker [1992] originally aimed to develop systems for functional programming without garbage collection.
% TODO: pervasive synonym
However, as noted by Wakeling and Runciman [1991], Wadler's effort relied greatly on pervasive copying.
The associated performance overhead of a reliance on copying was too large for use in real-world systems programming efforts.
Thus, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
The resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables [Baker 1995].
\cite{oxide}
% TODO: explain use-once concept in more detail with reference to Baker's paper

However, techniques like the 'use-once' variables [Baker 1995] differentiate themselves from Rust by requiring that all objects must be managed uniquely.
Instead, Rust allows the programmer to use unique references [Minsky 1996] with unguarded mutation or to use shared references without mutation.
This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration from work on ownership types and flexible alias protection [Clarke et al. 1998; Noble et al. 1998].

One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
% In Rust, like in owners-as-modifiers, an object can reference any other given object with read access.
The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
This allows for expressing many more programming idioms.

Furthermore, we can link certain owners-as-modifiers concepts to concepts from Rust.
Firstly, siblings...

% TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?

Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section from the Rust Book \cite{rust-book}.
If anything was retrieved from other sources, it would be cited as such.

% \subsection{From theory to Rust}
% \label{sec:use-once-variables}

% TODO: fix this section by reformatting the copied text and adding quotes from the references

% Rust's notion of ownership rests atop a long lineage of work, harkening back to the early days of linear logic [Girard 1987], and especially efforts by Wadler [1991] and Baker [1992] to develop systems for functional programming without garbage collection.
% However, as noted by Wakeling and Runciman [1991], Wadler’s effort relied greatly on pervasive copying.
% This reliance on copying and the associated performance penalty would not suffice for real world systems programming efforts, and thus, Rust’s ownership model is best understood as instead building off of Baker’s work on Linear Lisp where linearity enabled efficient reuse of objects in memory [Baker 1992, 1994a,b,1995].
% The resemblance is especially strong between Rust without borrowing and Baker’s ’use-once’ variables [Baker 1995].

% Rust’s main departure from techniques like ’use-once’ variables [Baker 1995] is a softening of a rather stringent requirement: namely, that everything must be managed uniquely.
% Instead, Rust allows the programmer to locally make a decision to use unique references [Minsky 1996] with unguarded mutation or to use shared references without such mutation.
% This flexibility in choosing arises at the point where the programmer creates a new reference, and draws inspiration fromwork on ownership types and flexible alias protection [Clarke et al. 1998; Noble et al. 1998].

% \subsection{How Rust implements ownership types}
% \label{sec:rust-ownership-types}

@@

@@

% \paragraph{Limits of the ownership approach}

@@

>>>>

\keywords{Ownership, Type, Safety, Rust}
\maketitle

@@

Memory safety refers to the state of a program where memory pointers or references always refer to valid memory.

@@

We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
The main problem here is that such memory issues cannot be detected at compile time.

@@

@@

The advantage of this model is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs. \cite{ownership-types-survey}
% TODO: which requirements for verficiation?

% \subsection{Limitations}
% \subsection{From ownership types to Rust}

% These early ownership type models carry some flaws with them. (lack features to be actually used)
% For example, in a system using the owners-as-dominators approach, one cannot programmatically create iterators.

% We can link certain owners-as-modifiers concepts to concepts from Rust.
% Firstly, siblings...

@@

Rust's notion of ownership is the culmination of a long line of work.
Many researchers have aimed to develop systems for functional programming without garbage collection.
According to Weiss et al. \cite{oxide}, it is best understood that Rust's ownership model is based off of Baker's work on Linear Lisp \cite{linear-lisp} where linearity enabled efficient reuse of objects in memory.
Furthermore, the resemblance is especially strong between Rust \emph{without} borrowing and Baker's 'use-once' variables.
'Use-once' variables are bound to linear (unshared, unaliased or singly-referenced) objects and are cheap to access and manage, because they require no synchronization or tracing garbage collection \cite{use-once}.

However, techniques like the 'use-once' variables differentiate themselves from Rust by requiring that all objects must be managed uniquely.
Instead, Rust allows the use of shared references without mutation or unique references with unguarded mutation \cite{oxide}.

One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
This allows for expressing more programming idioms.

@@

Most of the Rust-related information gathered in this chapter is derived from the "Understanding Ownership" section in the Rust Book \cite{rust-book}.

@@

% lifetime elision

% explain this example:
% A quick definition of ownership: you cannot have mutability and alias at the same time.
% My own example:

% \paragraph{Limits of Rust's ownership system}
% \paragraph{Limits of the ownership approach}

@@

% cpp biggest flaw: no compile time checks at all

% cpp comparison:
% http://nercury.github.io/c++/intro/2017/01/22/cpp-for-rust-devs.html
% https://www.tangramvision.com/blog/c-rust-interior-mutability-moving-and-ownership
